---
title: "Advanced Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
## Use ragg for better font rendering if available
if (requireNamespace("ragg", quietly = TRUE)) {
  options(summata.use_ragg = TRUE)
}

if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(
    dev = "ragg_png",
    fig.retina = 1,
    collapse = TRUE,
    comment = "##>",
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 5,
    out.width = "100%"
  )
} else {
  knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "##>",
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 5,
    out.width = "100%"
  )
}
options(width = 180)

## Only save figures when building pkgdown site (not for CRAN)
save_figures <- identical(Sys.getenv("IN_PKGDOWN"), "true")

## Use vignettes/figures for generated plots (pkgdown only)
fig_dir <- "figures"
if (save_figures && !dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

## Function for saving/displaying plots
## For pkgdown: saves PNG and returns path for include_graphics
## For CRAN: just prints the plot and returns empty string
save_example_plot <- function(plot, filename, width = 8, height = 5) {
  if (save_figures) {
    filepath <- file.path(fig_dir, filename)
    final_width <- attr(plot, "recommended_dims")$width %||% width
    final_height <- attr(plot, "recommended_dims")$height %||% height
    
    if (requireNamespace("ragg", quietly = TRUE)) {
      ## Use ragg directly - it uses 'res' not 'dpi', and defaults to 72
      ragg::agg_png(filepath, 
                    width = final_width, 
                    height = final_height, 
                    units = "in",
                    res = 72)
      print(plot)
      grDevices::dev.off()
    } else {
      ggplot2::ggsave(filepath, plot, 
                      width = final_width, 
                      height = final_height, 
                      dpi = 72)
    }
    return(filepath)
  } else {
    print(plot)
    return(NULL)
  }
}

## Wrapper for include_graphics that handles NULL (CRAN case)
show_plot <- function(plot, filename, width = 8, height = 5) {
  if (!save_figures) {
    return(invisible(NULL))  # Do not render anything for CRAN
  }
  path <- save_example_plot(plot, filename, width, height)
  if (!is.null(path)) {
    knitr::include_graphics(path)
  }
}
```

Standard regression analysis assumes independent observations and constant effects across subgroups. In practice, these assumptions often fail: observations may be clustered (e.g., subjects within study sites), effects may vary by subgroup (interaction), or the baseline hazard may differ across strata. This vignette demonstrates advanced analytical techniques to address these complexities.

For mathematical details on the methods covered here, see the [Statistical Foundations](statistical_foundations.html) article.

---

# Preliminaries

The examples in this vignette use the `clintrial` dataset included with `summata`:

```{r setup}
library(summata)
library(survival)
library(ggplot2)

data(clintrial)
data(clintrial_labels)

# Examine the clustering structure
table(clintrial$site)
```

The `clintrial` dataset includes 10 study sites, providing a natural clustering variable for hierarchical analysis.

---

# Interaction Effects

Interaction analysis tests whether the association between a predictor and outcome varies across levels of another variable. Interactions are specified using colon notation in the `interactions` parameter; main effects are automatically included alongside the interaction term.

## **Example 1:** Two-Way Interaction

Specify interactions using colon notation in the `interactions` parameter:

```{r}
example1 <- fit(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage"),
  interactions = c("sex:treatment"),
  model_type = "glm",
  labels = clintrial_labels
)

example1
```

## **Example 2:** Multiple Interactions

Test multiple effect modifiers simultaneously:

```{r}
example2 <- fit(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment", "stage"),
  interactions = c("sex:treatment", "stage:treatment"),
  model_type = "coxph",
  labels = clintrial_labels
)

example2
```

## **Example 3:** Continuous × Categorical Interaction

Test whether effects of a continuous variable vary by group:

```{r}
example3 <- fit(
  data = clintrial,
  outcome = "los_days",
  predictors = c("age", "sex", "treatment", "stage", "surgery"),
  interactions = c("age:treatment"),
  model_type = "lm",
  labels = clintrial_labels
)

example3
```

## **Example 4:** Interaction in fullfit()

Include interaction terms in combined univariable/multivariable analysis:

```{r}
example4 <- fullfit(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage", "sex:treatment"),
  model_type = "glm",
  method = "all",
  labels = clintrial_labels
)

example4
```

## **Example 5:** Testing Interaction Significance

Use `compfit()` to formally evaluate whether interactions improve model fit:

```{r}
example5 <- compfit(
  data = clintrial,
  outcome = "surgery",
  model_list = list(
    "Main Effects" = c("age", "sex", "treatment", "stage"),
    "Sex × Treatment" = c("age", "sex", "treatment", "stage"),
    "Stage × Treatment" = c("age", "sex", "treatment", "stage"),
    "Both Interactions" = c("age", "sex", "treatment", "stage")
  ),
  interactions_list = list(
    NULL,
    c("sex:treatment"),
    c("stage:treatment"),
    c("sex:treatment", "stage:treatment")
  ),
  model_type = "glm",
  labels = clintrial_labels
)

example5
```

## **Example 6:** Forest Plot with Interactions

Visualize models including interaction terms:

```{r}
interaction_model <- fit(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage"),
  interactions = c("sex:treatment"),
  model_type = "glm",
  labels = clintrial_labels
)

example6 <- glmforest(
  x = attr(interaction_model, "model"),
  title = "Logistic Regression with Interaction",
  labels = clintrial_labels,
  indent_groups = TRUE,
  zebra_stripes = TRUE
)
```

```{r, echo = FALSE, out.width = "100%"}
show_plot(example6, "adv_ex6.png")
```

---

# Mixed-Effects Models

Mixed-effects models (multilevel or hierarchical models) account for correlation within clusters by incorporating random effects. The `summata` package supports linear mixed models (`lmer`), generalized linear mixed models (`glmer`), and Cox mixed models (`coxme`).

Random effects are specified using lme4 syntax within the `predictors` parameter:

| Syntax | Meaning |
|:-------|:--------|
| `(1\|group)` | Random intercepts by group |
| `(x\|group)` | Random intercepts and slopes for *x* |
| `(1\|g1) + (1\|g2)` | Crossed random effects |

## **Example 7:** Random Intercepts (Linear Mixed)

The `(1|site)` syntax specifies a random intercept for each site:

```{r}
example7 <- fit(
  data = clintrial,
  outcome = "los_days",
  predictors = c("age", "sex", "treatment", "stage", "(1|site)"),
  model_type = "lmer",
  labels = clintrial_labels
)

example7
```

## **Example 8:** Random Intercepts (Logistic Mixed)

For binary outcomes with clustering:

```{r}
example8 <- fit(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage", "(1|site)"),
  model_type = "glmer",
  labels = clintrial_labels
)

example8
```

## **Example 9:** Random Slopes

Allow effects to vary by cluster:

```{r}
example9 <- fit(
  data = clintrial,
  outcome = "los_days",
  predictors = c("age", "sex", "treatment", "stage", "(1 + treatment|site)"),
  model_type = "lmer",
  labels = clintrial_labels
)

example9
```

## **Example 10:** Cox Mixed-Effects Model

For survival outcomes with clustering:

```{r}
example10 <- fit(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment", "stage", "(1|site)"),
  model_type = "coxme",
  labels = clintrial_labels
)

example10
```

## **Example 11:** Forest Plot from Mixed Model

Visualize fixed effects from mixed-effects models:

```{r}
example11 <- glmforest(
  x = attr(example8, "model"),
  title = "Logistic Mixed Model (Fixed Effects)",
  labels = clintrial_labels,
  indent_groups = TRUE,
  zebra_stripes = TRUE
)
```

```{r, echo = FALSE, out.width = "100%"}
show_plot(example11, "adv_ex11.png")
```

## **Example 12:** Comparing Random-Effects Specifications

Use `compfit()` to compare different random-effects structures:

```{r}
example12 <- compfit(
  data = clintrial,
  outcome = "los_days",
  model_list = list(
    "Random Intercepts" = c("age", "sex", "treatment", "stage", "(1|site)"),
    "Random Slopes" = c("age", "sex", "treatment", "stage", "(1 + treatment|site)")
  ),
  model_type = "lmer",
  labels = clintrial_labels
)

example12
```

---

# Stratification and Clustering

Cox models support stratification (separate baseline hazards) and clustered standard errors (robust variance estimation).

## **Example 13:** Stratified Cox Model

Stratification allows nonproportional hazards across strata without estimating stratum effects. Use the `strata` parameter:

```{r}
example13 <- fit(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment"),
  strata = "site",
  model_type = "coxph",
  labels = clintrial_labels
)

example13
```

## **Example 14:** Cluster-Robust SEs

For robust inference with correlated observations, cluster-robust standard errors account for within-cluster correlation. Use the `cluster` parameter:

```{r}
example14 <- fit(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment", "stage"),
  cluster = "site",
  model_type = "coxph",
  labels = clintrial_labels
)

example14
```

The following table summarizes when to use each approach:

| Approach | When to Use |
|:---------|:------------|
| Stratification | Proportional hazards assumption violated for a variable |
| Clustered SE | Correlation within clusters; robust inference |
| Mixed models | Model cluster-specific effects; prediction at cluster level |

---

# Weighted Regression

For survey data or inverse probability weighting, use the `weights` parameter. Weights can account for sampling design, non-response, or confounding adjustment.

## **Example 15:** Weighted Analysis

```{r}
# Create example weights
clintrial$analysis_weight <- runif(nrow(clintrial), 0.5, 2.0)

example15 <- fit(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage"),
  weights = "analysis_weight",
  model_type = "glm",
  labels = clintrial_labels
)

example15
```

---

# Advanced uniscreen() Features

The `uniscreen()` function supports advanced model specifications including random effects and stratification.

## **Example 16:** Univariable Screening with Random Effects

Screen multiple predictors while accounting for clustering:

```{r}
example16 <- uniscreen(
  data = clintrial,
  outcome = "surgery",
  predictors = c("age", "sex", "treatment", "stage"),
  model_type = "glmer",
  random = "(1|site)",
  labels = clintrial_labels
)

example16
```

## **Example 17:** Cox Screening with Mixed Effects

For survival outcomes with clustering:

```{r}
example17 <- uniscreen(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment", "stage", "ecog"),
  model_type = "coxme",
  random = "(1|site)",
  labels = clintrial_labels
)

example17
```

## **Example 18:** Forest Plot from Univariable Screening

The `uniforest()` function visualizes univariable screening results:

```{r}
uni_results <- uniscreen(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  predictors = c("age", "sex", "treatment", "stage", "ecog", "grade"),
  model_type = "coxph",
  labels = clintrial_labels
)

example18 <- uniforest(
  uni_results,
  title = "Univariable Screening Results",
  indent_groups = TRUE,
  zebra_stripes = TRUE
)
```

```{r, echo = FALSE, out.width = "100%"}
show_plot(example18, "adv_ex18.png")
```

---

# Advanced multifit() Features

The `multifit()` function supports interactions and mixed-effects models when testing a predictor across multiple outcomes.

## **Example 19:** Multi-Outcome with Interactions

Test effect modification across multiple outcomes:

```{r}
example19 <- multifit(
  data = clintrial,
  outcomes = c("surgery", "pfs_status", "os_status"),
  predictor = "treatment",
  covariates = c("age", "sex", "stage"),
  interactions = c("treatment:sex"),
  labels = clintrial_labels,
  parallel = FALSE
)

example19
```

## **Example 20:** Multi-Outcome with Mixed Effects

Account for clustering when testing effects across outcomes:

```{r}
example20 <- multifit(
  data = clintrial,
  outcomes = c("surgery", "pfs_status"),
  predictor = "treatment",
  covariates = c("age", "sex"),
  random = "(1|site)",
  model_type = "glmer",
  labels = clintrial_labels,
  parallel = FALSE
)

example20
```

## **Example 21:** Survival Multi-Outcome with Mixed Effects

For multiple survival outcomes with clustering:

```{r}
example21 <- multifit(
  data = clintrial,
  outcomes = c("Surv(pfs_months, pfs_status)",
               "Surv(os_months, os_status)"),
  predictor = "treatment",
  covariates = c("age", "sex"),
  random = "(1|site)",
  model_type = "coxme",
  labels = clintrial_labels,
  parallel = FALSE
)

example21
```

---

# Complete Workflow Example

The following demonstrates a complete advanced analysis workflow:

```{r, fig.width = 12, fig.height = 8}
# Step 1: Screen risk factors for primary outcome
risk_screening <- uniscreen(
  data = clintrial,
  outcome = "os_status",
  predictors = c("age", "sex", "bmi", "smoking", "diabetes",
                 "hypertension", "stage", "ecog", "treatment"),
  model_type = "glm",
  p_threshold = 0.20,
  labels = clintrial_labels
)

risk_screening

# Step 2: Test key exposure across multiple outcomes
effects <- multifit(
  data = clintrial,
  outcomes = c("surgery", "pfs_status", "os_status"),
  predictor = "treatment",
  covariates = c("age", "sex", "stage"),
  columns = "both",
  labels = clintrial_labels,
  parallel = FALSE
)

effects

# Step 3: Visualize effects
forest_plot <- multiforest(
  effects,
  title = "Effects Across Outcomes",
  indent_predictor = TRUE,
  zebra_stripes = TRUE
)
```

```{r, echo = FALSE, out.width = "100%"}
show_plot(forest_plot, "adv_workflow_forest.png")
```

---

# Best Practices

## Interaction Analysis

1. Include main effects for all variables in interactions
2. Test formally using interaction *p*-values
3. Non-significant interactions do not prove effect homogeneity
4. Consider domain plausibility when interpreting

## Mixed-Effects Models

1. Use when data exhibits natural clustering
2. Start with random intercepts; add random slopes if justified
3. Monitor convergence; simplify if necessary
4. Compare to fixed-effects models using `compfit()`

## Method Selection

| Scenario | Recommended Approach |
|:---------|:---------------------|
| Clustered observations | Mixed-effects models |
| Robust inference needed | Clustered standard errors |
| PH assumption violated | Stratification |
| Effect modification | Interaction terms |
| Survey data | Weighted regression |

## Sample Size Considerations

- **Interactions**: Require larger samples than main effects models
- **Mixed models**: Need sufficient clusters (≥20–30 recommended)
- **Stratification**: Each stratum needs adequate events

---

# Common Issues

## Convergence Problems in Mixed Models

Simplify the random effects structure if convergence fails:

```{r, eval = FALSE}
# Start with random intercepts only
fit(data, outcome, c(predictors, "(1|site)"), model_type = "lmer")

# Check for:
# - Few observations per cluster
# - Near-zero variance in random effects
# - Highly correlated predictors
```

## Interpreting Interactions

For categorical × categorical interactions, the coefficient represents the additional effect beyond the sum of main effects:

```{r, eval = FALSE}
# Access model for detailed interpretation
model <- attr(result, "model")
summary(model)
```

## Forest Plots with Interactions

Forest plots display all terms including interactions. For complex interaction patterns, consider stratified analyses or effect plots.

---

# Further Reading

- [Statistical Foundations](statistical_foundations.html): Mathematical details for all model types
- [Descriptive Tables](descriptive_tables.html): `desctable()` for baseline characteristics
- [Regression Modeling](regression_modeling.html): `fit()`, `uniscreen()`, and `fullfit()`
- [Model Comparison](model_comparison.html): `compfit()` for comparing models
- [Table Export](table_export.html): Export to PDF, Word, and other formats
- [Forest Plots](forest_plots.html): Visualization of regression results
- [Multivariate Analysis](multivariate_analysis.html): `multifit()` for multi-outcome analysis
