---
title: "Survival Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 150
)

# Use ragg for better font rendering if available
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}

options(width = 180)
```

Survival analysis requires specialized summary tables that report time-to-event outcomes in formats appropriate for longitudinal research. While `desctable()` includes basic survival summaries (median with 95% CI), detailed survival analysis often requires more comprehensive reporting: survival probabilities at specified time points, multiple quantiles, and group comparisons with appropriate statistical tests.
 
The `survtable()` function generates publication-ready survival tables with flexible output options. This vignette demonstrates its capabilities using the included sample dataset.

---

# Preliminaries

The examples in this vignette use the `clintrial` dataset included with `summata`:

```{r setup}
library(summata)
library(survival)

data(clintrial)
data(clintrial_labels)
```

The `clintrial` dataset contains `r nrow(clintrial)` observations with time-to-event variables suitable for demonstrating survival summaries.

---

# Basic Usage

## Function Signature

The `survtable()` function uses the familiar `Surv()` syntax for specifying survival outcomes:

```{r, eval = FALSE}
survtable(data, outcome, by, times, probs, ...)
```

where `data` is the dataset, `outcome` specifies the survival endpoint using `Surv()` notation, `by` defines the grouping variable, `times` specifies landmark time points, and `probs` specifies survival quantiles to report.

---

# Landmark Survival Estimates

Landmark survival estimates report the probability of survival at specific time points (*e.g.*, 1-year, 2-year survival rates).

## **Example 1:** Basic Landmark Analysis

The following example reports survival probabilities at 12, 24, and 36 months:

```{r}
example1 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24, 36),
  time_unit = "months"
)

example1
```

## **Example 2:** Combining Time Points and Median Survival

A common reporting format includes both landmark survival rates and median survival time:

```{r}
example2 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24),
  probs = 0.5,
  time_unit = "months",
  total = FALSE
)

example2
```

## **Example 3:** Remove Median Survival Column

By default, a column showing median survival is displayed (`probs = 0.5`). To remove it, set `probs = NULL`.

```{r}
example3 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24),
  probs = NULL,
  time_unit = "months",
  total = FALSE
)

example3
```

---

# Survival Quantiles

Survival quantiles report the time at which a specified proportion of subjects have experienced the event. The median survival time (50th percentile) is the most common, but other quantiles provide additional context.

## **Example 4:** Survival Quartiles

Report multiple survival quantiles by specifying the `probs` parameter:

```{r}
example4 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "stage",
  times = NULL,
  probs = c(0.25, 0.5, 0.75),
  labels = clintrial_labels
)

example4
```

---

# Multiple Endpoints

Studies often include multiple time-to-event outcomes, such as progression-free survival (PFS) and overall survival (OS). The `survtable()` function handles multiple endpoints in a single call.

## **Example 5:** PFS and OS Comparison

Pass a vector of outcomes to compare multiple survival endpoints:

```{r}
example5 <- survtable(
  data = clintrial,
  outcome = c("Surv(pfs_months, pfs_status)", "Surv(os_months, os_status)"),
  by = "treatment",
  times = c(12, 24),
  probs = 0.5,
  time_unit = "months",
  total = FALSE,
  labels = c(
    "Surv(pfs_months, pfs_status)" = "Progression-Free Survival",
    "Surv(os_months, os_status)" = "Overall Survival"
  )
)

example5
```

---

# Output Variations

## **Example 6:** Cumulative Event Rates

For competing risks analyses or when reporting event rates rather than survival probabilities, use `type = "risk"` to display cumulative incidence (1 − survival):

```{r}
example6 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24, 36),
  type = "risk",
  time_unit = "months"
)

example6
```

## **Example 7:** Including At-Risk Counts

The number at risk at each time point provides context for the precision of survival estimates:

```{r}
example7 <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24),
  stats = c("survival", "ci", "n_risk"),
  time_unit = "months",
  total = FALSE
)

example7
```

---

# Pairing with Kaplan-Meier Curves

Survival tables are often presented alongside Kaplan-Meier curves in publications. While `summata` focuses on tabular output, the `survminer` and `ggsurvfit` packages provide excellent options for survival curves.

## **Example 8:** Coordinated Figure-Table Output

The following workflow produces a matched figure-table pair suitable for publication:

```{r, eval = FALSE}
library(ggsurvfit)
library(survival)

# Fit the survival model
km_fit <- survfit(Surv(os_months, os_status) ~ treatment, data = clintrial)

# Create Kaplan-Meier plot with risk table
ggsurvfit(km_fit) +
  add_confidence_interval() +
  add_risktable() +
  add_quantile(y_value = 0.5, linetype = "dashed") +
  scale_ggsurvfit() +
  labs(
    title = "Overall Survival by Treatment",
    x = "Time (months)",
    y = "Survival Probability"
  ) +
  theme_minimal()

# Generate the companion table
surv_table <- survtable(
  data = clintrial,
  outcome = "Surv(os_months, os_status)",
  by = "treatment",
  times = c(12, 24, 36),
  probs = 0.5,
  time_unit = "months"
)

# Export both for publication
ggsave("km_curve.pdf", width = 8, height = 6)
table2pdf(surv_table, "survival_table.pdf", 
          caption = "Table 2. Survival Estimates by Treatment Group")
```

This workflow ensures that the Kaplan-Meier curve and survival table report consistent time points and groupings.

---

# Exporting Tables

Survival tables can be exported to various formats using the standard `summata` export functions. See the [Table Export](table_export.html) vignette for comprehensive documentation.

```{r, eval = FALSE}
# Microsoft Word
table2docx(
  table = example1,
  file = "SurvivalTable.docx",
  caption = "Table 2. Survival Estimates by Treatment Group"
)

# PDF (requires LaTeX)
table2pdf(
  table = example1,
  file = "SurvivalTable.pdf",
  caption = "Table 2. Survival Estimates by Treatment Group"
)
```

---

# Best Practices

## Time Point Selection

When selecting landmark time points, consider the following:

1. Use clinically meaningful intervals (e.g., 1-year, 2-year for oncology)
2. Ensure adequate follow-up at selected time points
3. Match time points across treatment arms for comparability
4. Consider the median follow-up when selecting the latest time point

## Reporting Recommendations

1. Always report confidence intervals alongside point estimates
2. Include the number at risk when space permits
3. Specify the time unit in table headers or footnotes
4. Use consistent decimal precision across estimates

---

# Further Reading

- [Descriptive Tables](descriptive_tables.html): `desctable()` for baseline characteristics
- [Regression Modeling](regression_modeling.html): Cox regression with `fit()` and `fullfit()`
- [Forest Plots](forest_plots.html): Visualization of hazard ratios
- [Table Export](table_export.html): Export to PDF, Word, and other formats
