---
title: "Statistical Foundations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

This article provides a comprehensive reference for the statistical methods implemented in `summata`. It serves as a technical companion to the package vignettes, detailing the mathematical foundations, assumptions, and interpretation of each supported model class.

For practical guidance on using these methods, see the [Regression Modeling](regression_modeling.html) vignette.

---

# Notation and Conventions

Throughout this article, we adopt the following notation:

| Symbol | Description |
|:-------|:------------|
| *Y* | Response (outcome) variable |
| *X*₁, ..., *X*ₚ | Predictor (covariate) variables |
| *β*₀ | Intercept |
| *β*₁, ..., *β*ₚ | Regression coefficients |
| *n* | Sample size |
| *i* | Observation index (*i* = 1, ..., *n*) |
| *j* | Cluster index (for hierarchical data) |
| *t* | Time (for survival models) |
| *g*(·) | Link function |
| exp(·) | Exponential function |
| log(·) | Natural logarithm |

---

# Linear Regression

## Model Specification

Linear regression models the expected value of a continuous response as a linear combination of predictors:

$$E[Y_i | X_i] = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip}$$

Equivalently, in matrix notation:

$$\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}$$

where $\boldsymbol{\varepsilon} \sim N(\mathbf{0}, \sigma^2\mathbf{I})$.

## Assumptions

1. **Linearity**: The relationship between predictors and response is linear
2. **Independence**: Observations are independent
3. **Homoscedasticity**: Constant error variance across all levels of predictors
4. **Normality**: Errors are normally distributed (for inference)
5. **No multicollinearity**: Predictors are not perfectly correlated

## Parameter Estimation

Coefficients are estimated by ordinary least squares (OLS):

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$$

## Interpretation

Each coefficient *β*ⱼ represents the expected change in *Y* for a one-unit increase in *X*ⱼ, holding all other predictors constant.

| Predictor Type | Interpretation of *β* |
|:---------------|:----------------------|
| Continuous | Change in *Y* per unit increase in *X* |
| Binary (0/1) | Difference in *Y* between groups |
| Categorical | Difference from reference category |

## Goodness of Fit

**Coefficient of determination (*R*²):**
$$R^2 = 1 - \frac{\sum_i (Y_i - \hat{Y}_i)^2}{\sum_i (Y_i - \bar{Y})^2}$$

**Adjusted *R*²:**
$$R^2_{\text{adj}} = 1 - \frac{(1 - R^2)(n - 1)}{n - p - 1}$$

**Root mean squared error (RMSE):**
$$\text{RMSE} = \sqrt{\frac{1}{n}\sum_i (Y_i - \hat{Y}_i)^2}$$

## Implementation

In `summata`, linear regression is specified with `model_type = "lm"`:

```{r}
fit(data, outcome = "continuous_var", predictors, model_type = "lm")
```

---

# Generalized Linear Models

Generalized linear models (GLMs) extend linear regression to accommodate non-normal response distributions through three components:

1. **Random component**: Distribution of *Y* from the exponential family
2. **Systematic component**: Linear predictor $\eta = \mathbf{X}\boldsymbol{\beta}$
3. **Link function**: $g(\mu) = \eta$, where $\mu = E[Y]$

## Logistic Regression

### Model Specification

For binary outcomes *Y* ∈ {0, 1}, logistic regression models the log-odds of the event:

$$\log\left(\frac{P(Y_i = 1 | X_i)}{1 - P(Y_i = 1 | X_i)}\right) = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$$

Equivalently:
$$P(Y_i = 1 | X_i) = \frac{\exp(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip})}{1 + \exp(\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip})}$$

### Assumptions

1. **Binary outcome**: *Y* takes values 0 or 1
2. **Independence**: Observations are independent
3. **Linearity in log-odds**: Log-odds is a linear function of predictors
4. **No multicollinearity**: Predictors are not perfectly correlated
5. **Large sample**: Adequate events per predictor (≥10 rule of thumb)

### Interpretation: Odds Ratios

The exponentiated coefficient exp(*β*ⱼ) yields the **odds ratio (OR)**:

$$\text{OR} = \exp(\beta_j) = \frac{\text{Odds}(Y = 1 | X_j + 1)}{\text{Odds}(Y = 1 | X_j)}$$

| OR Value | Interpretation |
|:---------|:---------------|
| OR = 1 | No association |
| OR > 1 | Increased odds of outcome |
| OR < 1 | Decreased odds of outcome |

For a continuous predictor, the OR represents the multiplicative change in odds per one-unit increase. For a categorical predictor, the OR compares each level to the reference category.

### Implementation

```{r}
fit(data, outcome = "binary_var", predictors, model_type = "glm", family = "binomial")
```

## Poisson Regression

### Model Specification

For count outcomes *Y* ∈ {0, 1, 2, ...}, Poisson regression models the log of the expected count:

$$\log(E[Y_i | X_i]) = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$$

### Assumptions

1. **Count outcome**: *Y* is a non-negative integer
2. **Independence**: Observations are independent
3. **Equidispersion**: Mean equals variance (E[*Y*] = Var(*Y*))
4. **Log-linear relationship**: Log of expected count is linear in predictors

### Interpretation: Rate Ratios

The exponentiated coefficient exp(*β*ⱼ) yields the **rate ratio (RR)**:

$$\text{RR} = \exp(\beta_j) = \frac{E[Y | X_j + 1]}{E[Y | X_j]}$$

### Overdispersion

When Var(*Y*) > E[*Y*] (overdispersion), standard Poisson SEs are underestimated. Solutions include:

- **Quasipoisson**: Estimates dispersion parameter *φ*
- **Negative binomial**: Models additional variance component

### Implementation

```{r}
# Standard Poisson
fit(data, outcome = "count_var", predictors, model_type = "glm", family = "poisson")

# Overdispersed counts
fit(data, outcome = "count_var", predictors, model_type = "glm", family = "quasipoisson")
```

## Negative Binomial Regression

### Model Specification

Negative binomial regression extends Poisson regression with an additional dispersion parameter *θ*:

$$Y_i \sim \text{NegBin}(\mu_i, \theta)$$
$$\log(\mu_i) = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$$

The variance is:
$$\text{Var}(Y_i) = \mu_i + \frac{\mu_i^2}{\theta}$$

As *θ* → ∞, the negative binomial converges to Poisson.

### When to Use

- Residual deviance >> residual degrees of freedom in Poisson model
- Variance substantially exceeds mean
- Overdispersion test is significant

### Implementation

```{r}
fit(data, outcome = "count_var", predictors, model_type = "negbin")
```

Requires the **MASS** package.

## Gamma Regression

### Model Specification

For positive, continuous, right-skewed outcomes, Gamma regression models:

$$Y_i \sim \text{Gamma}(\mu_i, \phi)$$

With log link (default):
$$\log(E[Y_i | X_i]) = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$$

### Assumptions

1. **Positive outcome**: *Y* > 0
2. **Gamma distribution**: Right-skewed, variance proportional to mean squared
3. **Independence**: Observations are independent

### Interpretation

With log link, exp(*β*ⱼ) represents the **multiplicative effect** (ratio):

$$\frac{E[Y | X_j + 1]}{E[Y | X_j]} = \exp(\beta_j)$$

### Common Applications

- Healthcare costs
- Length of hospital stay
- Biomarker concentrations
- Time durations

### Implementation

```{r}
fit(data, outcome = "positive_continuous", predictors, model_type = "glm", family = "Gamma")
```

## Quasi-Likelihood Models

### Quasibinomial

For overdispersed binary data, quasibinomial regression estimates a dispersion parameter:

$$\text{Var}(Y_i) = \phi \cdot \mu_i(1 - \mu_i)$$

where *φ* > 1 indicates overdispersion. Coefficients are identical to binomial GLM, but standard errors are adjusted.

### Quasipoisson

For overdispersed count data:

$$\text{Var}(Y_i) = \phi \cdot \mu_i$$

### Implementation

```{r}
# Overdispersed binary
fit(data, outcome, predictors, model_type = "glm", family = "quasibinomial")

# Overdispersed counts
fit(data, outcome, predictors, model_type = "glm", family = "quasipoisson")
```

## Summary of GLM Families

| Family | Outcome Type | Link | Effect Measure |
|:-------|:-------------|:-----|:---------------|
| `binomial` | Binary | logit | Odds ratio |
| `quasibinomial` | Binary (overdispersed) | logit | Odds ratio |
| `poisson` | Count | log | Rate ratio |
| `quasipoisson` | Count (overdispersed) | log | Rate ratio |
| `gaussian` | Continuous | identity | Coefficient |
| `Gamma` | Positive continuous | log | Ratio |
| `inverse.gaussian` | Positive continuous | 1/μ² | — |

---

# Survival Analysis

## Cox Proportional Hazards Model

### Model Specification

The Cox model relates the hazard function to predictors:

$$h(t | X_i) = h_0(t) \exp(\beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip})$$

where:

- *h*(*t* | *X*) is the hazard (instantaneous risk) at time *t*
- *h*₀(*t*) is the baseline hazard (unspecified)
- exp(*β*ⱼ) is the hazard ratio for predictor *j*

### Assumptions

1. **Proportional hazards**: Hazard ratios are constant over time
2. **Independence**: Observations are independent (or appropriately modeled)
3. **Censoring**: Non-informative censoring
4. **No multicollinearity**: Predictors are not perfectly correlated

### Interpretation: Hazard Ratios

The **hazard ratio (HR)** = exp(*β*ⱼ) represents the relative hazard:

$$\text{HR} = \frac{h(t | X_j + 1)}{h(t | X_j)} = \exp(\beta_j)$$

| HR Value | Interpretation |
|:---------|:---------------|
| HR = 1 | No association with survival |
| HR > 1 | Increased hazard (shorter survival) |
| HR < 1 | Decreased hazard (longer survival) |

### Goodness of Fit

**Concordance (Harrell's C-index):**
$$C = P(\hat{h}_i > \hat{h}_j | T_i < T_j)$$

Proportion of concordant pairs; C = 0.5 indicates no discrimination, C = 1 indicates perfect discrimination.

### Testing Proportional Hazards

The proportional hazards assumption can be tested using:

- Scaled Schoenfeld residuals
- Log-log survival plots
- Time-dependent covariates

### Implementation

```{r}
fit(data, outcome = "Surv(time, status)", predictors, model_type = "coxph")
```

Requires the **survival** package.

## Conditional Logistic Regression

### Model Specification

For matched case-control studies, conditional logistic regression conditions on the stratum:

$$\log\left(\frac{P(Y_i = 1 | X_i, \text{stratum})}{P(Y_i = 0 | X_i, \text{stratum})}\right) = \beta_1 X_{i1} + \cdots + \beta_p X_{ip}$$

Note: No intercept is estimated; it is absorbed by the stratum.

### When to Use

- 1:1 or 1:*m* matched case-control designs
- Stratified analyses where stratum-specific effects are nuisance parameters

### Implementation

```{r}
fit(data, outcome = "case_status", predictors, model_type = "clogit", strata = "match_id")
```

Requires the **survival** package.

---

# Mixed-Effects Models

Mixed-effects models (hierarchical or multilevel models) account for correlation within clusters by incorporating random effects.

## General Framework

For observation *i* within cluster *j*:

$$g(E[Y_{ij} | X_{ij}, u_j]) = \mathbf{X}_{ij}\boldsymbol{\beta} + \mathbf{Z}_{ij}\mathbf{u}_j$$

where:

- **β** are fixed effects (population-averaged)
- **u**ⱼ ~ N(**0**, **Σ**) are random effects (cluster-specific deviations)
- **Z**ᵢⱼ is the random effects design matrix

## Linear Mixed-Effects Models

### Model Specification

For continuous outcomes:

$$Y_{ij} = \beta_0 + \beta_1 X_{1ij} + \cdots + \beta_p X_{pij} + u_{0j} + u_{1j}X_{1ij} + \varepsilon_{ij}$$

where:

- *u*₀ⱼ ~ N(0, σ²ᵤ₀) is the random intercept for cluster *j*
- *u*₁ⱼ ~ N(0, σ²ᵤ₁) is the random slope (if included)
- εᵢⱼ ~ N(0, σ²) is the residual error

### Random Effects Syntax

| Syntax | Meaning |
|:-------|:--------|
| `(1\|group)` | Random intercepts by group |
| `(x\|group)` | Random intercepts and slopes for *x* |
| `(1\|group1) + (1\|group2)` | Crossed random effects |
| `(1\|group1/group2)` | Nested random effects |

### Implementation

```{r}
fit(data, outcome, c(predictors, "(1|cluster)"), model_type = "lmer")
```

Requires the **lme4** package.

## Generalized Linear Mixed-Effects Models

### Model Specification

GLMMs extend mixed-effects models to non-normal outcomes:

$$g(E[Y_{ij} | X_{ij}, u_j]) = \beta_0 + \beta_1 X_{1ij} + \cdots + \beta_p X_{pij} + u_{0j}$$

Common families:

- **Binomial**: Clustered binary outcomes
- **Poisson**: Clustered count outcomes

### Interpretation

Fixed effects are interpreted as **conditional** (cluster-specific) effects. The population-averaged effect may differ due to the non-linear link function.

### Implementation

```{r}
# Binary outcome with clustering
fit(data, outcome, c(predictors, "(1|cluster)"), model_type = "glmer", family = "binomial")

# Count outcome with clustering
fit(data, outcome, c(predictors, "(1|cluster)"), model_type = "glmer", family = "poisson")
```

Requires the **lme4** package.

## Cox Mixed-Effects Models

### Model Specification

For clustered survival data:

$$h_{ij}(t) = h_0(t) \exp(\beta_1 X_{1ij} + \cdots + \beta_p X_{pij} + u_j)$$

where *u*ⱼ ~ N(0, σ²ᵤ) is the cluster-specific frailty term.

### When to Use

- Patients nested within hospitals or physicians
- Repeated events within subjects
- Multi-center clinical trials

### Implementation

```{r}
fit(data, outcome = "Surv(time, status)", c(predictors, "(1|cluster)"), model_type = "coxme")
```

Requires the **coxme** package.

## Sample Size Considerations

Mixed-effects models require adequate sample sizes at both levels:

| Level | Recommendation |
|:------|:---------------|
| Clusters | ≥20–30 for stable variance estimates |
| Observations per cluster | ≥5 for random intercepts |
| Total observations | Standard regression guidelines apply |

---

# Interaction Effects

## Model Specification

An interaction occurs when the effect of one predictor depends on the level of another. For predictors *X* and *Z*:

$$g(E[Y]) = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 (X \times Z)$$

where:

- *β*₁ is the effect of *X* when *Z* = 0
- *β*₂ is the effect of *Z* when *X* = 0
- *β*₃ is the interaction effect (modification of *X* effect by *Z*)

## Interpretation

### Continuous × Continuous

For continuous *X* and *Z*, the effect of *X* on *Y* is:

$$\frac{\partial E[Y]}{\partial X} = \beta_1 + \beta_3 Z$$

The effect of *X* varies linearly with *Z*.

### Continuous × Categorical

For continuous *X* and binary *Z* (0/1):

- When *Z* = 0: Effect of *X* is *β*₁
- When *Z* = 1: Effect of *X* is *β*₁ + *β*₃

### Categorical × Categorical

For two categorical variables, each combination of levels has a distinct effect. The interaction coefficient represents the additional effect beyond the sum of main effects.

## Testing Interactions

1. **Wald test**: Test H₀: *β*₃ = 0
2. **Likelihood ratio test**: Compare models with and without interaction
3. **Model comparison**: Use `compfit()` to compare AIC/BIC

## Implementation

```{r}
fit(data, outcome, predictors, interactions = c("X:Z"), model_type = "glm")
```

---

# Model Quality Metrics

## Information Criteria

### Akaike Information Criterion (AIC)

$$\text{AIC} = -2\log L + 2k$$

where *L* is the maximized likelihood and *k* is the number of parameters. AIC balances fit and complexity; lower values indicate better models.

### Bayesian Information Criterion (BIC)

$$\text{BIC} = -2\log L + k\log n$$

BIC penalizes complexity more heavily than AIC, especially for large samples.

### Interpretation

| Comparison | Guideline |
|:-----------|:----------|
| ΔAIC < 2 | Models essentially equivalent |
| ΔAIC 2–7 | Some support for lower-AIC model |
| ΔAIC > 10 | Strong support for lower-AIC model |

## Discrimination Metrics

### Concordance (C-statistic)

For binary outcomes or survival models:

$$C = P(\hat{p}_i > \hat{p}_j | Y_i = 1, Y_j = 0)$$

| C Value | Discrimination |
|:--------|:---------------|
| 0.5 | No discrimination (random) |
| 0.6–0.7 | Poor |
| 0.7–0.8 | Acceptable |
| 0.8–0.9 | Excellent |
| > 0.9 | Outstanding |

### Brier Score

For binary outcomes:

$$\text{Brier} = \frac{1}{n}\sum_i (Y_i - \hat{p}_i)^2$$

Lower values indicate better calibration; Brier = 0 is perfect.

## Explained Variation

### R² (Linear Models)

$$R^2 = 1 - \frac{SS_{\text{res}}}{SS_{\text{tot}}}$$

Proportion of variance explained.

### Pseudo-R² (GLMs)

**McFadden's pseudo-R²:**
$$R^2_{\text{McF}} = 1 - \frac{\log L_{\text{full}}}{\log L_{\text{null}}}$$

**Nagelkerke's pseudo-R²:**
$$R^2_N = \frac{1 - (L_{\text{null}}/L_{\text{full}})^{2/n}}{1 - L_{\text{null}}^{2/n}}$$

Values are not directly comparable to linear *R*²; 0.2–0.4 typically indicates good fit for logistic models.

---

# Composite Model Score (CMS)

## Rationale

When comparing multiple candidate models, researchers typically examine several quality metrics: information criteria (AIC, BIC), discrimination (concordance), calibration (Brier score), and explained variation (*R*² or pseudo-*R*²). Each metric captures a different aspect of model performance, and no single measure is universally optimal for model selection.

The **Composite Model Score (CMS)** addresses this challenge by combining multiple metrics into a single summary value, facilitating rapid comparison across candidate models. This approach:

1. Synthesizes complementary information from diverse metrics
2. Provides model-type-specific weighting based on statistical conventions
3. Normalizes metrics to a common 0–100 scale for interpretability
4. Accounts for convergence issues that may invalidate model results

The CMS is intended as a screening tool to identify promising models, not as a replacement for substantive evaluation. Final model selection should always consider domain knowledge, parsimony, and the specific research objectives.

## Methodology

### Normalization

Each component metric is transformed to a 0–100 scale where higher values indicate better performance:

**Information criteria** (AIC, BIC) — Lower is better:
$$S_{\text{AIC}} = 100 \times \left(1 - \frac{\text{AIC}_i - \text{AIC}_{\min}}{\text{AIC}_{\max} - \text{AIC}_{\min}}\right)$$

**Concordance** (*C*-statistic) — Higher is better, with 0.5 as floor:
$$S_C = \begin{cases} 
0 & C \leq 0.5 \\
200 \times (C - 0.5) & C > 0.5
\end{cases}$$

**Pseudo-*R*²** (McFadden) — Scaled recognizing typical range:
$$S_{R^2} = \min(100, \; 250 \times R^2_{\text{McF}})$$

This scaling reflects that McFadden's *R*² rarely exceeds 0.4 even for well-fitting models.

**Brier score** — Lower is better, with 0.25 as no-skill baseline:
$$S_{\text{Brier}} = 100 \times \left(1 - \frac{\text{Brier}}{0.25}\right)$$

**Convergence** — Categorical assessment:
$$S_{\text{conv}} = \begin{cases}
100 & \text{converged normally} \\
70 & \text{suspect (warnings)} \\
30 & \text{failed to converge}
\end{cases}$$

### Weighting

Component scores are combined using model-type-specific weights that reflect the relative importance of each metric for that model class:

| Model Type | Convergence | AIC | Concordance | Pseudo-*R*² | Brier | Global *p* |
|:-----------|:-----------:|:---:|:-----------:|:-----------:|:-----:|:----------:|
| GLM (logistic) | 0.15 | 0.25 | 0.40 | 0.15 | 0.05 | — |
| Cox PH | 0.15 | 0.30 | 0.40 | — | — | 0.15 |
| Linear | 0.15 | 0.25 | — | 0.45 | — | 0.15 |

For mixed-effects models, additional components include marginal *R*², conditional *R*², and ICC.

### Final Score

The CMS is computed as a weighted sum:

$$\text{CMS} = \sum_j w_j \times S_j$$

where *w*ⱼ are the weights and *S*ⱼ are the normalized component scores.

## Interpretation

| CMS Range | Interpretation |
|:----------|:---------------|
| 85–100 | Excellent model quality |
| 75–84 | Very good |
| 65–74 | Good |
| 55–64 | Fair |
| < 55 | Poor |

These thresholds provide general guidance. A model with CMS = 80 is not necessarily "better" than one with CMS = 78—small differences should prompt examination of individual metrics and substantive considerations.

## Limitations

1. **Relative ranking**: CMS values are relative to the set of models being compared; they do not indicate absolute model quality
2. **Weight sensitivity**: Results depend on the chosen weights; users with strong priors about metric importance can supply custom weights
3. **Metric availability**: Not all metrics are available for all model types; weights are renormalized when components are missing
4. **No substitute for judgment**: The CMS facilitates screening but does not replace careful model diagnostics and domain expertise

## Implementation

```{r}
# Compare candidate models
comparison <- compfit(
  data = mydata,
  outcome = "y",
  predictors = list(
    "Model 1" = c("x1", "x2"),
    "Model 2" = c("x1", "x2", "x3"),
    "Model 3" = c("x1", "x2", "x3", "x4")
  ),
  model_type = "logistic"
)

# Custom weights emphasizing discrimination
comparison_custom <- compfit(
  data = mydata,
  outcome = "y",
  predictors = model_list,
  model_type = "logistic",
  scoring_weights = list(
    convergence = 0.10,
    aic = 0.20,
    concordance = 0.50,
    pseudo_r2 = 0.15,
    brier = 0.05
  )
)
```

---

# Confidence Intervals

## Wald Confidence Intervals

For coefficient *β* with standard error SE(*β*):

$$\beta \pm z_{1-\alpha/2} \cdot \text{SE}(\beta)$$

where *z*₁₋α/₂ is the standard normal quantile (1.96 for 95% CI).

## Profile Likelihood Intervals

More accurate for small samples or boundary estimates; based on the likelihood ratio:

$$\{\beta : 2[\log L(\hat{\beta}) - \log L(\beta)] \leq \chi^2_{1,1-\alpha}\}$$

## Confidence Intervals for Effect Measures

For exponentiated coefficients (OR, HR, RR):

$$\exp\left(\beta \pm z_{1-\alpha/2} \cdot \text{SE}(\beta)\right)$$

---

# Hypothesis Testing

## Wald Test

Tests H₀: *β* = 0 using:

$$z = \frac{\hat{\beta}}{\text{SE}(\hat{\beta})}$$

Under H₀, *z* ~ N(0, 1).

## Likelihood Ratio Test

Compares nested models:

$$\Lambda = -2[\log L_{\text{reduced}} - \log L_{\text{full}}]$$

Under H₀, Λ ~ χ²ₖ, where *k* is the difference in parameters.

## Global Test

Tests H₀: all *β*ⱼ = 0 (except intercept):

- **Linear models**: F-test
- **GLMs**: Likelihood ratio test
- **Cox models**: Likelihood ratio, Wald, or score test

---

# References

## Methodological

- Agresti, A. (2013). *Categorical Data Analysis* (3rd ed.). Wiley.
- Harrell, F. E. (2015). *Regression Modeling Strategies* (2nd ed.). Springer.
- Hosmer, D. W., Lemeshow, S., & Sturdivant, R. X. (2013). *Applied Logistic Regression* (3rd ed.). Wiley.
- Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data*. Springer.
- Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.

## Software

- R Core Team (2024). *R: A Language and Environment for Statistical Computing*. https://www.R-project.org/
- Therneau, T. (2024). *survival: Survival Analysis*. R package. https://CRAN.R-project.org/package=survival
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, 67(1), 1–48.
- Therneau, T. (2024). *coxme: Cox Mixed-Effects Models*. R package. https://CRAN.R-project.org/package=coxme
- Venables, W. N., & Ripley, B. D. (2002). *Modern Applied Statistics with S* (4th ed.). Springer. (MASS package)

---

# See Also

- [Regression Modeling](regression_modeling.html): Practical examples using `fit()`, `uniscreen()`, and `fullfit()`
- [Advanced Workflows](advanced_workflows.html): Interactions and mixed-effects models
- [Model Comparison](model_comparison.html): Comparing alternative specifications with `compfit()`
- [Forest Plots](forest_plots.html): Visualization of regression results
